---
title: "ESM 211 - Homework 1"
author: "Laura Ingulsrud"
date: "January 17, 2020"
output:
  html_document:
    df_print: paged
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, include=FALSE}

# Load necessary packages
library(tidyverse)

```

&nbsp;
&nbsp;
####1. 

###**Eureka dune grass**

####2.

```{r swalleniadata, include=FALSE}

# Read in data for Eureka dune grass (Swallenia alexandrae) on Saline Dune
swallenia <- read.csv("Swallenia.csv")

# Use two-sample t-test to assess whether and how mean abundance changed from 2009 to 2010
swallenia_t <- t.test(swallenia$count_2009, swallenia$count_2010, var.equal = TRUE)

swallenia_t

```

A significance level of 0.05 was chosen for $\alpha$ because it indicates a 5% risk of concluding a difference in mean abundance exists, when there actually is no difference in mean abundance of Eureka dune grass between the two years. There is a 0.05 probability that a Type I error will be committed (the null hypothesis is true, and we reject it).

Eureka dune grass mean abundance on Saline Dune between 2009 and 2010 did not differ significantly (p = `r round(swallenia_t$p.value, 2)`, $\alpha$ = 0.05). 

####3.

```{r}

# Do a paired t-test
swallenia_tpaired <- t.test(swallenia$count_2009, swallenia$count_2010, paired = TRUE)

swallenia_tpaired

```

The true difference in means between Eureka dune grass abundance on Saline Dune between 2009 and 2010 is not equal to 0; therefore, the change in abundance differed significantly (p = `r round(swallenia_tpaired$p.value, 2)`, $\alpha$ = 0.05).

####4. The paired t-test (second analyses) is more appropriate because there is a one-to-one relationship between values in the two Eureka dune grass samples from 2009 and 2010. Since plants were counted in 11 quadrats across the dune, the counts were from the same locations between 2009 and 2010. Since Eureka dune grass spreads outward from itself, it likely exhibits clumped dispersion, and so the quadrats can be expected to capture plant abundance year to year.

####5. 


###**Yellowstone grizzly bears**

####6. 

```{r grizzlydata}

# Read in grizzly data
grizzlydata <- read.csv("grizzlydata.csv")

# Did the grizzly bear population decline over the period of 1959 - 1968?

# Filter grizzly data to only include years 1959 - 1968
grizzly1959_1968 <- grizzlydata %>% 
  filter(between(Year, 1959, 1968))
  
grizzly1959_1968_summarize <- grizzly1959_1968 %>% 
  summarize(
    mean1959_1968 = round(mean(N), 2), # Mean = 43.6
    sd1959_1968 = sd(mean(N), 2), # SD = NA
    samplesize1959_1968 = length(N), # Sample size = 10
  )

# Create exploratory histogram
grizzly1959_1968_hist <- ggplot(grizzly1959_1968, aes(x = N)) + 
  geom_histogram(bins = 10) +
    scale_x_continuous(expand = c(0,0)) +
    scale_y_continuous(expand = c(0,0)) +
    theme_classic() +
    theme(text = element_text(family = "Times New Roman")) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
    theme(legend.position="none") +
    theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"))

grizzly1959_1968_hist

# Create exploratory quantile quantile plot
grizzly1959_1968_qq <- ggplot(grizzly1959_1968, aes(sample = N)) + 
  geom_qq() +
  theme_classic() +
  theme(legend.position="none") +
  theme(text = element_text(family = "Times New Roman")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

grizzly1959_1968_qq

# Create exploratory scatterplot
gg_grizzly1959_1968 <- ggplot(grizzly1959_1968, aes(x = Year, y = N)) +
  geom_point()

gg_grizzly1959_1968
# Overall, it looks like the trend is generally a negative linear relationship. However, it seems the population was relatively stable from 1959 to 1965, but it drastically declined after 1965.

# Perform linear regression to describe the trend in grizzly bear population from 1959 - 1968
grizzly1959_1968_model <- lm(N ~ Year, data = grizzly1959_1968)

grizzly1959_1968_model
# Grizzly population = 1543 - 0.76(Year)

# View model diagnostics
par(mfrow = c(2,2))
plot(grizzly1959_1968_model)

# Explore model fit and signifiance
summary(grizzly1959_1968_model)

```


####7. 

```{r}

# Filter grizzly data to only include years 1969 - 1978
grizzly1969_1978 <- grizzlydata %>% 
  filter(between(Year, 1969, 1978))

# Create exploratory scatterplot
gg_grizzly1969_1978 <- ggplot(grizzly1969_1978, aes(x = Year, y = N)) +
  geom_point()

gg_grizzly1969_1978
# Overall, it looks like the trend is generally a negative linear relationship. However, there are some outliers on both ends.

# Perform linear regression to describe the trend in grizzly bear population from 1959 - 1968
grizzly1969_1978_model <- lm(N ~ Year, data = grizzly1969_1978)

grizzly1969_1978_model
# Grizzly population = 1532.38 - 0.76(Year)

```


####8. 

```{r}

# Filter grizzly data to only include years 1978 - 1997
grizzly1978_1997 <- grizzlydata %>% 
  filter(between(Year, 1978, 1997))

# Create exploratory scatterplot
gg_grizzly1978_1997 <- ggplot(grizzly1978_1997, aes(x = Year, y = N)) +
  geom_point()

gg_grizzly1978_1997
# Overall, it looks like the trend is generally a positive linear relationship.

# Perform linear regression to describe the trend in grizzly bear population from 1959 - 1968
grizzly1978_1997_model <- lm(N ~ Year, data = grizzly1978_1997)

grizzly1978_1997_model
# Grizzly population = -6043.13 + 3.07(Year)

```


####9. 