---
title: "ESM 211 - Homework 1"
author: "Laura Ingulsrud"
date: "January 17, 2020"
output:
  html_document:
    df_print: paged
    theme: cosmo
---

<style>
h1.title{
    text-align: center
  }
h4.author{
    text-align: center
  }
h4.date{
    text-align: center
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, include=FALSE}

# Load necessary packages
library(tidyverse)

```

&nbsp;
&nbsp;

####1. Give three reasons why a plant or animal might be patchily distributed like this. For each, describe whether you would expect the location of the high-density patches to be consistent from year to year.

This figure shows a pattern of clumped dispersion, in which individuals of a population are clustered in groups.

1. The high-density patches could be in areas of suitable habitat. For an animal population, the high-density patches might be locations of abundant food or safety from predators. For a plant population, the high-density patches might be or areas of optimal soil nutrients, water, and sunlight. Depending on how suitable habitat patches increase, decrease, or move spatially, the species would be expected to follow this pattern.

2. For plants, it could be a plant species that spreads outward from itself, or drops its seeds straight to the ground, rather than its seeds being dispersed more widely. This type of plant species would remain in clusters. In this situation, the location of these patches would be expected to be consistent from year to year.

3. For animals, it could be an animal species that lives in groups, such as herds of elephants. For this situation, whether the location of these patches remains consistent from year to year depends on the reason for the animals groups’ locations. For example, if they are grouped in areas of high food availability, location constancy depends on whether those areas remain rich in food or not.

&nbsp;

##Eureka dune grass

####2. Read in the data, and use a two-sample t-test (e.g., using t.test() in R) to assess whether and how mean abundance changed from 2009 to 2010. Choose a level for α, and justify your choice. On the basis of this analysis, what do you conclude about the change in the grass’s abundance?

```{r swalleniadata}

# Read in data for Eureka dune grass (Swallenia alexandrae) on Saline Dune
swallenia <- read.csv("Swallenia.csv")

# Use two-sample t-test to assess whether and how mean abundance changed from 2009 to 2010
swallenia_t <- t.test(swallenia$count_2009, swallenia$count_2010, var.equal = TRUE)

swallenia_t

```

A significance level of 0.05 was chosen for $\alpha$ because it indicates a 5% risk of concluding a difference in mean abundance exists, when there actually is no difference in mean abundance of Eureka dune grass between the two years. There is a 0.05 probability that a Type I error will be committed (the null hypothesis is true, and we reject it). A significance level of 0.05 is common in hypothesis testing.

Eureka dune grass mean abundance on Saline Dune between 2009 and 2010 did not differ significantly (p = `r round(swallenia_t$p.value, 2)`, $\alpha$ = 0.05). 

&nbsp;

####3. Because the plant was counted in the same plots each year, you can also do a paired t-test (R hint: use paired = TRUE in t.test()). On the basis of this analysis, what do you conclude about the change in the grass’s abundance?

```{r}

# Do a paired t-test
swallenia_tpaired <- t.test(swallenia$count_2009, swallenia$count_2010, paired = TRUE)

swallenia_tpaired

```

The true difference in means between Eureka dune grass abundance on Saline Dune between 2009 and 2010 is not equal to 0; therefore, the change in abundance differed significantly (p = `r round(swallenia_tpaired$p.value, 2)`, $\alpha$ = 0.05).

&nbsp;

####4. Which of these analyses is more appropriate? Why? (Hint: think about your answers to question 1, and whether any apply here).

The paired t-test (second analyses) is more appropriate because there is a one-to-one relationship between values in the two Eureka dune grass samples from 2009 and 2010. Since plants were counted in 11 quadrats across the dune, the counts were from the same locations between 2009 and 2010. Since Eureka dune grass spreads outward from itself, it likely exhibits clumped dispersion, and so the quadrats can be expected to capture plant abundance year to year.

&nbsp;

####5. Write a short paragraph to the park superintendent describing your finding about the changes (if any) to the dune grass population.

&nbsp;

###Yellowstone grizzly bears

####6. Open the file, and look at the data from 1959 to 1968. Did the grizzly bear population decline over this period? Support your conclusion with graphs, statistics, and logical reasoning.

```{r grizzlydata}

# Read in grizzly data
grizzlydata <- read.csv("grizzlydata.csv")

# Did the grizzly bear population decline over the period of 1959 - 1968?

# Filter grizzly data to only include years 1959 - 1968
grizzly1959_1968 <- grizzlydata %>% 
  filter(between(Year, 1959, 1968))
  
grizzly1959_1968_summarize <- grizzly1959_1968 %>% 
  summarize(
    mean1959_1968 = round(mean(N), 2), # Mean = 43.6
    sd1959_1968 = sd(mean(N), 2), # SD = NA
    samplesize1959_1968 = length(N), # Sample size = 10
  )

# Create exploratory histogram
grizzly1959_1968_hist <- ggplot(grizzly1959_1968, aes(x = N)) + 
  geom_histogram(bins = 10) +
    scale_x_continuous(expand = c(0,0)) +
    scale_y_continuous(expand = c(0,0)) +
    theme_classic() +
    theme(text = element_text(family = "Times New Roman")) +
    theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
    theme(legend.position="none") +
    theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "black"))

grizzly1959_1968_hist

# Create exploratory quantile quantile plot
grizzly1959_1968_qq <- ggplot(grizzly1959_1968, aes(sample = N)) + 
  geom_qq() +
  theme_classic() +
  theme(legend.position="none") +
  theme(text = element_text(family = "Times New Roman")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

grizzly1959_1968_qq

# Create exploratory scatterplot
gg_grizzly1959_1968 <- ggplot(grizzly1959_1968, aes(x = Year, y = N)) +
  geom_point()

gg_grizzly1959_1968
# Overall, it looks like the trend is generally a negative linear relationship. However, it seems the population was relatively stable from 1959 to 1965, but it drastically declined after 1965.

# Perform linear regression to describe the trend in grizzly bear population from 1959 - 1968
grizzly1959_1968_model <- lm(N ~ Year, data = grizzly1959_1968)

grizzly1959_1968_model
# y = 1543 - 0.76x
# Grizzly population = 1543 - 0.76(Year)

# View model diagnostics
par(mfrow = c(2,2))
plot(grizzly1959_1968_model)

# Explore model fit and signifiance
summary(grizzly1959_1968_model)

# Prepare final graph of the linear regression model with the original data, model, and 95% confidence interval for predictions
grizzly1959_1968_graph <- ggplot(grizzly1959_1968, aes(x = Year, y = N)) +
  geom_point() +
  geom_smooth(method = lm, se = TRUE, size = 0.5, color = "gray20") +
  theme_bw() +
  scale_x_continuous(expand = c(0,0), limits = c(1958, 1969)) +
  labs(x = "Year", y = "Yellowstone Grizzly Bear Population")

grizzly1959_1968_graph

```

&nbsp;

####7. Select the data from 1969 to 1978. Did the population continue to decline? Was the decline faster than the period prior to the dump closures?

```{r}

# Filter grizzly data to only include years 1969 - 1978
grizzly1969_1978 <- grizzlydata %>% 
  filter(between(Year, 1969, 1978))

# Create exploratory scatterplot
gg_grizzly1969_1978 <- ggplot(grizzly1969_1978, aes(x = Year, y = N)) +
  geom_point()

gg_grizzly1969_1978
# Overall, it looks like the trend is generally a negative linear relationship. However, there are some outliers on both ends.

# Perform linear regression to describe the trend in grizzly bear population from 1959 - 1968
grizzly1969_1978_model <- lm(N ~ Year, data = grizzly1969_1978)

grizzly1969_1978_model
# Grizzly population = 1532.38 - 0.76(Year)

# Prepare final graph of the linear regression model with the original data, model, and 95% confidence interval for predictions
grizzly1969_1978_graph <- ggplot(grizzly1969_1978, aes(x = Year, y = N)) +
  geom_point() +
  geom_smooth(method = lm, se = TRUE, size = 0.5, color = "gray20") +
  theme_bw() +
  scale_x_continuous(expand = c(0,0), limits = c(1968, 1979)) +
  labs(x = "Year", y = "Yellowstone Grizzly Bear Population")

grizzly1969_1978_graph

```

&nbsp;

####8. Finally look at the data after 1978. Did population size continue to change? What was the direction and magnitude of the trend?

```{r}

# Filter grizzly data to only include years 1978 - 1997
grizzly1978_1997 <- grizzlydata %>% 
  filter(between(Year, 1978, 1997))

# Create exploratory scatterplot
gg_grizzly1978_1997 <- ggplot(grizzly1978_1997, aes(x = Year, y = N)) +
  geom_point()

gg_grizzly1978_1997
# Overall, it looks like the trend is generally a positive linear relationship.

# Perform linear regression to describe the trend in grizzly bear population from 1959 - 1968
grizzly1978_1997_model <- lm(N ~ Year, data = grizzly1978_1997)

grizzly1978_1997_model
# Grizzly population = -6043.13 + 3.07(Year)

# Prepare final graph of the linear regression model with the original data, model, and 95% confidence interval for predictions
grizzly1978_1997_graph <- ggplot(grizzly1978_1997, aes(x = Year, y = N)) +
  geom_point() +
  geom_smooth(method = lm, se = TRUE, size = 0.5, color = "gray20") +
  theme_bw() +
  scale_x_continuous(expand = c(0,0), limits = c(1977, 1998)) +
  labs(x = "Year", y = "Yellowstone Grizzly Bear Population")

grizzly1978_1997_graph

```

&nbsp;

####9. Write a short paragraph to the park superintendent describing your conclusions on the effects (if any) of the dump closures on the grizzly bear population.